---
title: "hw06: Data Wrangling"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In this assignment I mainly explored some basic functions of stringr which are common but useful in many cases, and performed split-apply-combine strategy with the `gapminder` dataset. 

A quick navigation:

- [Character data](#character-data)
  - [Paste](#paste)
  - [Combination](#combination)
  - [Extraction](#extraction)
  - [Wrapping](#wrapping)
  - [Trim and pad](#trim-and-pad)
  
- [Work with a nested data frame](#work-with-a-nested-data-frame)
  - [Split by nesting](#split-by-nesting)
  - [Apply a function](#apply-a-function)
  - [Combine and simplyfy](#combine-and-simplyfy)
  - [Explore the residuals](#explore-the-residuals)

```{r}
library(stringr)
library(ggplot2)
library(gapminder)
suppressPackageStartupMessages(library(tidyverse))
```


## Character data

### Paste

>1. In code that doesn’t use stringr, you’ll often see `paste()` and` paste0()`. What’s the difference between the two functions? What stringr function are they equivalent to? How do the functions differ in their handling of `NA`?

Try to combine the strings "Hello" and "world".

```{r}
paste("Hello","world")
paste("Hello","world",sep = "@")
paste0("Hello","world")
```

From the above, we could see that `paste()` function adds separate character (default with a space " ") when combining two strings. Whereas `paste0()` function combines strings without space. `str_c()` is kind of like a cross between these two functions. You could use it directly or use the `sep` argument to control how the strings are separated:

```{r}
str_c("Hello","world") # like `paste0`
str_c("Hello","world",sep = "@") # like `paste`
```

What if one string is `NA` here?
```{r}
x <- c("abc", NA)
str_c("|-", x, "-|")
paste("|-", x, "-|")
```
As shown above, both `str_c()` and `paste()` are vectorised, but the missing value `NA` would not be operated in `str_c()`.  

### Combination

>2. In your own words, describe the difference between the `sep` and `collapse` arguments to `str_c()`.

If there are two or more arguments of string, choose `sep`. If the strings are stored in a vector like the following example, use `collapse` instead.

```{r}
str_c(c("Hello", "world"), collapse = "@")
```

### Extraction

>3. Use `str_length()` and `str_sub()` to extract the middle character from a string. What will you do if the string has an even number of characters?

Let's define a function that extract the middle character from a string no matter its length is odd or even. By the trick of `ceiling`, we could find the median of the length.

```{r}
str_mid_char <- function(x){
  str_sub(x, ceiling(str_length(x)/2), ceiling((str_length(x)+1)/2))
}
str_mid_char("apple") # odd number of characters
str_mid_char("banana") # even number of characters
```

### Wrapping

>4. What does `str_wrap()` do? When might you want to use it?

`str_wrap()` allows the control of output format, i.e., wrapping strings into nicely formatted paragraphs, with the arguments of `width`, `indent` and `extent`. Inspired by the reply from our TA in [one post of github discussion](https://github.com/STAT545-UBC/Discussion-Internal/issues/61#issuecomment-435173377), I decided to explore how this function works with a long axis label when visulization. Just make a silly example as below. The second label decreases the readability and aesthetic of the axis because of the length of string. By the help with `str_wrap()`, the problem of format is solved.  

```{r}
long_label <- "very, very, very, very, very long label"
fixed_label <- str_wrap(long_label, width = 20)
df = data.frame(x = c("label", long_label, fixed_label), y = c(10, 15, 20)) # compare two strings in one axis 

ggplot(df, aes(x, y)) + 
  xlab("") + ylab("Number") +
  geom_bar(stat = "identity")
```


### Trim and pad

>5. What does str_trim() do? What’s the opposite of str_trim()?

`str_trim()` is used to remove whitespace from start and end of string. For example:
```{r}
(a <- "   Hello world\t")

str_trim(a)
```
Just notice that the space between the words is kept.

`str_pad()` works in an opposite way from this function.

```{r}
(a <- "Hello world")

str_pad(a, width = 20, side = "both")
```
Of course, you could change the padding character with the argument `pad`.

## Work with a nested data frame

This section is inspired by the [split-apply-combine lesson from Jenny Bryan](http://stat545.com/block024_group-nest-split-map.html). It refers to get apply a computation to all the pieces of a dataset and glue the results back together. In this way we could compute or graph these results more concisely. By this strategy, I tried to figure out the countries whose life expectancy has a sudden and large deviation from the general trend over years.

### Split by nesting

Nesting can be used to support more complex group-wise computation. When you nest data, the non-grouping variables are packaged into group-specific data frames that are held in a special variable called a list-column. I didn't use `kable()` since I wanted to show the structure of list-column.

```{r}
(gap_nested <- gapminder %>% 
  group_by(continent, country) %>% 
  nest()) # nest the data by country and continent

```

### Apply a function

The objective is to find the countries with outliers in life expectancy. So firstly write a function that fit a model of life expectancy against year, and return the residuals to detect countries where the model is a terrible fit.

```{r}
le_vs_yr <- function(df) {
  lm(lifeExp ~ I(year - 1950), data = df)$residuals # return residuals of the fitted model 
}

```

Use `purrr::map()` to apply the fitting function `le_vs_yr()` to all the countries and store the results into new list-column with `mutate()`.

```{r}
(gap_res <- gap_nested %>% 
  mutate(res = map(data, le_vs_yr)) %>% # fit the model with every country
  select(continent, country, res)) 
```

### Combine and simplyfy

The last step is now to simplify, preferably back to a normal tibble. We do this by retaining variables that are amenable to simplification and using `unnest()`, thus completing the circle.

```{r}
gap_res <- unnest(gap_res)

gap_res %>% 
  sample_n(10) %>% 
  knitr::kable()
```

### Explore the residuals

Finally we could have a look on the residuals of fitted models. Through the efforts above, `gap_res` is handy for numerical summarization and visualization.

```{r}
summary(gap_res$res)
```

The numerical summaries look reasonable. The points from the first quartile to the third quartile are well fit in the model. We conclude with a look at the full distribution.

```{r}
ggplot(gap_res, aes(x = res)) +
  geom_density() + 
  geom_rug() +
  ggtitle("The distribution of residuals") +
  theme_bw()
```

As shown in the graph, most points locate between -10 and 10. Let's find the countries with far outliers.

```{r}
gap_res %>% 
  filter(abs(res)>10) %>% # filter the outliers
  knitr::kable()
```

We can conclude that Cambodi, Lesotho, Rwanda, Swaziland and Zimbabwe have seen a huge fluctuation in their trend of life expectancy. Take Zimbabwe as an example. 

```{r}
gapminder %>% 
  filter(country == "Zimbabwe") %>% 
  ggplot(aes(x = year-1950, y = lifeExp)) +
  geom_smooth(se = FALSE, method = lm) + # fit linear regression
  geom_point()
```

